Spatial variation of the native colon microbiota in healthy adults

### 2018-01-30 submission/manuscript.Rmd
reviewer revisions completed

The human colon is an ecosystem comprised of numerous microenvironments that select for different microbiota. Concentrations of oxygen, water, and anti-microbial peptides change along the gut axis and influence which microbiota reside in each location. Microenvironments differ not only longitudinally along the colon, but also radially from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbiota and communities are important for understanding the etiology of complex diseases such as Colorectal Cancer (CRC) and Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or accelerated by perturbations the structure of the gut microbiota [@Yamauchi2012; @Forbes2016; @Halfvarson2017]. The manifestations of these diseases are known to vary based upon the location in which they occur. For instance, CRC that arises in the distal (left) colon are of hindgut origin and tend to have large chromosomal alterations indicative of chromosomal instability @Yamauchi2012. In contrast, CRC arising in the proximal (right) colon tumors are of midgut origin and tend to be sessile and microsatellite instable (MSI with BRAF and KRAS mutations) @Yamauchi2012. In addition to the environmental gradients within the colon, the distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively [@Yamauchi2012; @Albenberg2014; @Donaldson2015]. In IBD patients, disease occurring in the distal colon extending proximally is usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur anywhere along the GI tract, most commonly in the ileum and the cecum @Forbes2016. UC presents as continuous disease with only mucosal involvement, where as CD has skip lesions and full thickness involvement that may cause abscesses, strictures and fistulas @Forbes2016. Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbiota and their metabolites likely vary as well, and may influence the heterogeneous disease prognoses of IBD and CRC. Because CRC can be a long-term complication of IBD, the distribution of microbiota is important to understanding the pathophysiology of both diseases.

@@

The few existing profiles of the microbial spatial variation of the colon have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy or surgery @Donaldson2015. While the latter method allows investigators to acquire samples from inside the human colon, typically these procedures are preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel. Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer [@Jalanka2014; @Harrell2012]. As such, what little information we do have about the spatial distribution of the microbiota in the proximal and distal colon is confounded by the bowel preparation procedure.

@@

### Methods
#### Human subjects
The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number HUM00082721. Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study. Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl, Versed and Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive. 20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure. There were 13 female and 7 male subjects ranging in age from 25 to 64.

#### Sample collection
At a baseline visit, subjects gave consent and were given a home collection stool kit (Zymo). At least one week prior to the scheduled colonoscopy, subjects collected whole stool at home and shipped the samples to a research coordinator on ice. Notably, subjects did not undergo any bowel preparation method prior to sampling. On the procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System. Subjects were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate. A flexible sigmoidoscope was first inserted about 25cm into the colon and jumbo biopsy forceps used to collect the luminal contents. Two luminal samples were collected and the contents immediately deposited into RNAlater (Fisher) and flash-frozen in liquid nitrogen. The forceps were withdrawn and new biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter. Three mucosal biopsies were collected and flash-frozen in RNAlater. These samples comprised the distal colon samples. The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the proximal colon. Samples were then collected in the same manner as was done in the distal colon and the colonoscope withdrawn. All samples were stored at -80$^\circ$C.

#### Sample processing, sequencing and analysis
DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples. The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013. Sequences were curated using the mothur software as described previously @Schloss2009. The sequences were assigned a taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff. Sequencing and analysis of a mock community revealed the error rate to be 0.018%. Samples were rarefied to 4231 sequences per sample in order to reduce the effects of uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator and $\theta$~YC~ calculator metrics in mothur version 1.39.5 @Schloss2009. $\theta$~YC~ distances were calculated to determine the dissimilarity between two samples. Random Forest classification models were built using the AUCRF R package using a leave-one-subject out approach @Calle_2011. For each model the data were split into a 19-subject training set and a 1-subject test set. The model was built and cross-validated using AUCRFcv on the training set. The model was then tested on the left-out patient. This process was repeated iteratively for all subjects and results plotted as Reciever Operator Characteristic curves using the pROC R package @Robin_2011. Resultant models were used to identify the OTUs that were most important for classifying each location. Species-level information for sequences of interest was obtained by aligning the sequences to the GenBank nucleotide databse using blastn. The species name was only used if the identity score was $\ge$ 99% over the full-length of the contig and matched a single reference.

@@

When examining the relative abundance of the dominant phyla at each site (i.e. _Bacteriodes_ and _Firmicutes_), there was a wide amount of variation. This likely reflects not only the variability between human subjects, casued by differences in age, sex, diet, but also spatial "patchiness" in the gut microbiome. One study noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other @Hong2011. Similar patchiness was also observed in luminal contents and fecal samples themselves; there was separation of different interacting microbes along the length of a stool sample, for instance @Stearns2011. That said, across our samples, the mucosal samples harbor more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to luminal content in humans @Albenberg2014. Hence, the conclusions we were able to draw from phyla analysis may have been impacted by inter-subject patchiness.

@@

In addition to identifying features that are specific to each side of the gut, the ability of the Random Forest to classify samples can serve as a proxy for similarity. That is, a higher AUC value indicates the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value. For instance, the model separating the proximal and distal mucosa had an AUC of 0.850 whereas the model for classifying the proximal and distal lumen had a much lower AUC of 0.580. Further, the latter model required 44 OTUs to best separate the samples whereas the models separating the mucosa only needed 10 OTUs. The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested. We speculate that the model was less effective at classifying the proximal and distal luminal contents because the mucosal microenvironments have variable selective pressure along the colon than the luminal microenvironments.

@@

Spatial organization of _Fusobacteria_ and other bacterial species into polymicrobial biofilms that can invade the gut mucosa have been linked CRC (cite new paper). The organization into biofilms promotes tumorigenesis by allowing bacteria to grow near the epithelium, promoting inflammation, genotoxicity and metabolic changes that favor tumor cell growth (cite new paper and kaitlin). In one study of CRC biofilms and tumors, all of the proximal tumors examined contained a polymicrobial biofilm on the tumor mucosa @Dejea2014. Further examination of these tumors identified _Fusobacteria_ species as members of the biofilm. These results indicate that it is not only the presence of the bacterium but the tumor community as a whole that contributes to tumorigenesis @Dejea2014. A 'driver-passenger' model has been proposed as a mechanism for biofilm assembly in the gut (Tjalsma 2012, flynn). In this model, 'driver' species such as _Fusobacterium spp_ and _Porphyromonas spp_ exert tumorigenic effects locally and create a niche for colonization or adherence of 'passenger' species that comprise the rest of the biofilm (Tjalsma and Flynn). Thus the the distribution of these disease-associated microbes in healthy patients is of interest as their presence can be predictive of disease prior to the onset of symptoms @Baxter2016. A better understanding of the early microbial changes in the gut microboiome is essential for elucidating a mechanism for development of CRC or IBD subtypes in the proximal or distal colon.

@@

By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypotheses about how specific host-microbe interactions can affect disease progression of proximal and distal CRC and IBD subtypes. Further investigation of these samples using metagenomics and metatranscriptomics would illuminate the microbial activities in these gut microenvironments and could enhance microbiome-based screening and treatment modalities for colon disease.

>>>>

The human colon is an ecosystem comprised of numerous microenvironments that select for different microbiota. Concentrations of oxygen, water, and anti-microbial peptides change along the gut axis and influence which microbiota reside in each location. Microenvironments differ not only longitudinally along the colon, but also radially from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbiota and communities are important for understanding the etiology of complex diseases such as Colorectal Cancer (CRC) and Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or accelerated by perturbations of the structure of the gut microbiota [@Yamauchi2012; @Forbes2016; @Halfvarson2017]. The manifestations of these diseases are known to vary based upon the location in which they occur. For instance, CRC that arises in the distal (left) colon are of hindgut origin and tend to have large chromosomal alterations indicative of chromosomal instability @Yamauchi2012. In contrast, CRC arising in the proximal (right) colon are of midgut origin and tend to be sessile and microsatellite instable (MSI with BRAF and KRAS mutations) @Yamauchi2012. In addition to the environmental gradients within the colon, the distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively [@Yamauchi2012; @Albenberg2014; @Donaldson2015]. In IBD patients, disease occurring in the distal colon extending proximally is usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur anywhere along the GI tract, most commonly in the ileum and the cecum @Forbes2016. UC presents as continuous disease with only mucosal involvement, where as CD has skip lesions and full thickness involvement that may cause abscesses, strictures and fistulas @Forbes2016. Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbiota and their metabolites likely vary as well, and may influence the heterogeneous disease prognoses of IBD and CRC. Because CRC can be a long-term complication of IBD, the distribution of microbiota is important to understanding the pathophysiology of both diseases.

@@

The few existing profiles of the microbial spatial variation of the colon have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy or surgery @Donaldson2015. While invasive methods allow investigators to acquire samples from inside the human colon, typically these procedures are preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel. Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer [@Jalanka2014; @Harrell2012]. As such, what little information we do have about the spatial distribution of the microbiota in the proximal and distal colon is confounded by the bowel preparation procedure.

@@

### Methods
#### Human subjects
The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number HUM00082721. Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study. Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl, Versed and Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive. 20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure. There were 13 female and 7 male subjects ranging in age from 25 to 64. 18 of the 20 subjects had not used antibiotics within a year prior to the collection date and 2 had not used antibiotics within 6 months. None of the subjects had medical conditions requiring frequent or extended antibiotic use.

#### Sample collection
At a baseline visit, subjects gave consent and were given a home collection stool kit (Zymo). One to seven days prior to the scheduled colonoscopy, subjects collected whole stool at home and shipped the samples to a research coordinator on ice. Notably, subjects did not undergo any bowel preparation method prior to sampling. On the procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System. Subjects were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate. A flexible sigmoidoscope was first inserted about 25cm into the colon and jumbo biopsy forceps used to collect the luminal contents. Two luminal samples were collected and the contents immediately deposited into RNAlater (Fisher) and flash-frozen in liquid nitrogen. The forceps were withdrawn and new biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter. Three mucosal biopsies were collected and flash-frozen in RNAlater. These samples comprised the distal colon samples. The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the proximal colon. The proximal samples were taken from the ascending colon proximal to the hepatic flexure at 75-120cm depending on the subject. Samples were then collected in the same manner as was done in the distal colon and the colonoscope withdrawn. All samples were stored at -80$^\circ$C.

#### Sample processing, sequencing and analysis
DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples. The resulting DNA was normalized to equal concentrations across all samples and used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013. Sequences were curated using the mothur software as described previously @Schloss2009. The sequences were assigned a taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff. Sequencing and analysis of a mock community revealed the error rate to be 0.018%. Samples were rarefied to 4231 sequences per sample in order to reduce the effects of uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator and $\theta$~YC~ calculator metrics in mothur version 1.39.5 @Schloss2009. $\theta$~YC~ distances were calculated to determine the dissimilarity between two samples. Random Forest classification models were built using the AUCRF R package using a leave-one-subject out approach @Calle_2011. The Random Forest models were built using the full, non-rarefied, dataset as input. For each model the data were split into a 19-subject training set and a 1-subject test set. The model was built and cross-validated using 10-fold k cross-validation (AUCRFcv) on the training set to estimate the prediction error of the model. The resultant model was then used to predict the outcome the left-out subject. This process was repeated iteratively for all subjects and results plotted as Reciever Operator Characteristic curves using the pROC R package @Robin_2011. Resultant models were used to identify the OTUs that were most important for classifying each location. Species-level information for sequences of interest was obtained by aligning the sequences to the GenBank nucleotide databse using blastn. The species name was only used if the identity score was $\ge$ 99% over the full-length of the contig and matched a single reference.

@@

When examining the relative abundance of the dominant phyla at each site (i.e. _Bacteriodes_ and _Firmicutes_), there was a wide amount of variation. This likely reflects not only the variability between human subjects, caused by differences in age, sex, and diet, but may also reflect spatial patchiness in the gut microbiome within a subject. Patchiness refers to inconsistent distribution of microbial populations due to fluctuations in local resources @Rougharden1977. One study noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other @Hong2011. Similar patchiness was also observed in luminal contents and fecal samples themselves; there was separation of different interacting microbes along the length of a stool sample, for instance @Stearns2011. A third study that sampled six mucosal sites along the colon observed such patchiness in two of the three study subjects @Eckburg2005. While our subjects were not sampled frequently enough to draw specific conclusions about patchiness along the unprepped colon, we did observe some specific differences in mucosal versus luminal samples at the phylum level. The mucosal samples harbored more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to luminal content in humans @Albenberg2014. However, we must still consider that the results from phyla analysis may have been impacted by inter-subject patchiness.

@@

In addition to identifying features that are specific to each side of the gut, the ability of the Random Forest to classify samples can serve as a proxy for similarity. That is, a higher AUC value indicates the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value. For instance, the model separating the proximal and distal mucosa had an AUC of 0.850 whereas the model for classifying the proximal and distal lumen had a much lower AUC of 0.580. Further, the latter model required 44 OTUs to best separate the samples whereas the models separating the mucosa only needed 10 OTUs. The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested. We speculate that the model was less effective at classifying the proximal and distal luminal contents because the mucosal microenvironments have more variable selective pressure along the colon than the luminal microenvironments.

@@

Spatial organization of _Fusobacteria_ and other bacterial species into polymicrobial biofilms that can invade the gut mucosa have been linked to CRC @Li2017. The biofilms promote tumorigenesis by allowing bacteria to grow near the epithelium, inducing inflammation, genotoxicity and metabolic changes that favor tumor cell growth @Li2017. In one study of CRC biofilms and tumors, all of the proximal tumors examined contained a polymicrobial biofilm on the tumor mucosa @Dejea2014. Further examination of these tumors identified _Fusobacteria_ species as members of the biofilm. These results indicate that it is not only the presence of the bacterium but the tumor community as a whole that contributes to tumorigenesis @Dejea2014. A 'driver-passenger' model has been proposed as a mechanism for biofilm assembly in the gut [@Tjalsma2012; @Flynn2016]. In this model, 'driver' species such as _Fusobacterium spp_ and _Porphyromonas spp_ exert tumorigenic effects locally and create a niche for adherence of 'passenger' species that comprise the rest of the biofilm [@Tjalsma2012; @Flynn2016]. Thus the the distribution of these disease-associated microbes in healthy patients is of interest as their presence can be predictive of disease prior to the onset of symptoms @Baxter2016. A better understanding of the early microbial changes in the gut microboiome is essential for elucidating a mechanism for development of CRC or IBD subtypes in the proximal or distal colon.

@@

By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypotheses about how specific host-microbe interactions can affect disease progression of proximal and distal CRC and IBD subtypes. Future investigation of these samples using metagenomics and metatranscriptomics would illuminate the microbial activities in these gut microenvironments. Further, combining this approach with a more comprehensive sampling strategy along the unprepped colon could enhance microbiome-based screening and treatment modalities for colon disease.
